# survival-features
_Last modified `r format(Sys.time(), "%I:%M %p on %b %d, %Y")`. This document, R session image, version history, knitr cache, figures, and other associated datasets are located in `/inside/grotto/blin/trna-markers/luad/predict/`._

```{r setup, echo=FALSE, warning=FALSE, results=FALSE, message=FALSE, errors=FALSE, cache=FALSE}
library(knitr)
opts_chunk$set(cache = TRUE, cache.path = "/inside/grotto/blin/trna-markers/luad/predict/cache/survival-features/", fig.path = "/inside/grotto/blin/trna-markers/luad/predict/figure/survival-features/", eval = TRUE, echo = TRUE, warning = FALSE, results = FALSE, message = FALSE, autodep = TRUE, dev = "png", dpi = 300)
```

```{r libraries, cache=FALSE}
library(glmnet)
library(plyr)
library(glmnet)
library(ROCR)
library(GenomicRanges)
library(stringr)
library(reshape2)
library(survival)
set.seed(12)
source('/inside/home/blin/lib/R/ggsurv.R')
source('predict.R')
load('/inside/home/blin/grotto/data/hg19-srnas.RData')
load('/inside/home/blin/grotto/trna-markers/process-reads/luad-counts.RData')
load('coefs.RData')
luad_clinical <- luad_clinical[match(colnames(luad_adjusted_counts), luad_clinical$barcode), ]
ntrials <- 200
```

We perform feature selection and build a new model to classify patients into low/high risk groups.

## tsRNAs

### Build model, get coefficients

```{r tsrna-1}
metadata <- luad_clinical[luad_clinical$sample_type == "TP" & luad_clinical$days_survived > 0 & luad_clinical$days_survived < 3000 & !is.na(luad_clinical$days_survived), ]
counts <- luad_adjusted_counts[rownames(luad_adjusted_counts) %in% tsrna_coefs[tsrna_coefs$score > 1, ]$name, colnames(luad_adjusted_counts) %in% metadata$barcode]
metadata$dummy <- as.factor(1)
dataset <- setupTrainingTestingSets(metadata, "dummy", counts)
training_counts <- t(dataset$training_counts)
training_response <- Surv(dataset$training_metadata$days_survived, dataset$training_metadata$vital_status)
cv.glmnet(training_counts, training_response, family = "cox")
plot(model)
coef(model, s = model$lambda.min)
plot(model$glmnet.fit, label = TRUE)
```

### Naive Cox model prediction using summed hazard ratios

```{r tsrna-2}
# evaluate on early tumors - that's what's clinically relevant ;)
testing_metadata <- dataset$testing_metadata[dataset$testing_metadata$stage %in% c("Stage I", "Stage IA"), ]
testing_counts <- t(dataset$testing_counts[, colnames(dataset$testing_counts) %in% testing_metadata$barcode])
testing_response <- as.factor(ifelse(testing_metadata$days_survived >= 1095, ">3", "<3")) # test 3 year survival rate
pred <- predict(model, newx = testing_counts, s = model$lambda.min) 
prob <- prediction(-pred, testing_response) # flip signs because a positive hazard ratio gives a higher risk of death
perf <- performance(prob, "tpr", "fpr")
roc <- data.frame(TPR = unlist(perf@y.values), FPR = unlist(perf@x.values))
auc <- unlist(performance(prob, "auc")@y.values)
auc <- round(auc, 3)
ggplot(roc) + geom_line(aes(x = FPR, y = TPR)) + geom_abline(intercept = 0, slope = 1, colour = "gray") + ylab("TPR") + xlab("FPR")
auc
```

### Classify into high/low risk groups

```{r tsrna-3}
metadata <- dataset$testing_metadata
metadata$risk <- as.factor(ifelse(pred > median(pred), "High-risk", "Low-risk"))
km <- survfit(Surv(days_survived, vital_status) ~ risk, data = metadata)
ggsurv(km) + theme_bw() + ggtitle("Kaplan-Meier plot for risk based on tsRNA signature")
cox <- coxph(Surv(days_survived, vital_status) ~ risk, data = metadata)
summary(cox)
```

### Build new logistic regression model using hazard ratios


```{r tsrna-4}


```

## miRNAs

### Build model, get coefficients

```{r mirna-1}
metadata <- luad_clinical[luad_clinical$sample_type == "TP" & luad_clinical$days_survived > 0 & luad_clinical$days_survived < 3000 & !is.na(luad_clinical$days_survived), ]
counts <- luad_adjusted_counts[rownames(luad_adjusted_counts) %in% mirna_coefs[mirna_coefs$score > 1, ]$name, colnames(luad_adjusted_counts) %in% metadata$barcode]
tsrna_model <- buildModelGetCoefs(metadata, counts)
plot(model)
coef(model, s = model$lambda.min)
plot(model$glmnet.fit, label = TRUE)
```

### Naive Cox model prediction using summed hazard ratios

```{r glmnet-cox-2}
testing_counts <- t(dataset$testing_counts)
testing_response <- as.factor(ifelse(dataset$testing_metadata$days_survived >= 1095, ">3", "<3")) # test 1 year survival rate
pred <- predict(model, newx = testing_counts, s = model$lambda.min) 
prob <- prediction(-pred, testing_response) # flip signs because a positive hazard ratio gives a higher risk of death
perf <- performance(prob, "tpr", "fpr")
roc <- data.frame(TPR = unlist(perf@y.values), FPR = unlist(perf@x.values))
auc <- unlist(performance(prob, "auc")@y.values)
auc <- round(auc, 3)
ggplot(roc) + geom_line(aes(x = FPR, y = TPR)) + geom_abline(intercept = 0, slope = 1, colour = "gray") + ylab("TPR") + xlab("FPR")
auc
```

### Classify into high/low risk groups

```{r glmnet-cox-3}
metadata <- dataset$testing_metadata
metadata$risk <- as.factor(ifelse(pred > median(pred), "High-risk", "Low-risk"))
km <- survfit(Surv(days_survived, vital_status) ~ risk, data = metadata)
ggsurv(km) + theme_bw() + ggtitle("Kaplan-Meier plot for risk based on miRNA signature")
cox <- coxph(Surv(days_survived, vital_status) ~ risk, data = metadata)
summary(cox)
```

### Build new logistic regression model using hazard ratios


```{r glmnet-cox-4}


```

## Combined tsRNAs + miRNAs

### Build model, get coefficients

```{r mirna-1}
metadata <- luad_clinical[luad_clinical$sample_type == "TP" & luad_clinical$days_survived > 0 & luad_clinical$days_survived < 3000 & !is.na(luad_clinical$days_survived), ]
counts <- luad_adjusted_counts[rownames(luad_adjusted_counts) %in% c(mirna_coefs[mirna_coefs$score > 1, ]$name, tsrna_coefs[tsrna_coefs$score > 1, ]$name), colnames(luad_adjusted_counts) %in% metadata$barcode]
tsrna_model <- buildModelGetCoefs(metadata, counts)
plot(model)
coef(model, s = model$lambda.min)
plot(model$glmnet.fit, label = TRUE)
```

### Naive Cox model prediction using summed hazard ratios

```{r glmnet-cox-2}
testing_counts <- t(dataset$testing_counts)
testing_response <- as.factor(ifelse(dataset$testing_metadata$days_survived >= 1095, ">3", "<3")) # test 1 year survival rate
pred <- predict(model, newx = testing_counts, s = model$lambda.min) 
prob <- prediction(-pred, testing_response) # flip signs because a positive hazard ratio gives a higher risk of death
perf <- performance(prob, "tpr", "fpr")
roc <- data.frame(TPR = unlist(perf@y.values), FPR = unlist(perf@x.values))
auc <- unlist(performance(prob, "auc")@y.values)
auc <- round(auc, 3)
ggplot(roc) + geom_line(aes(x = FPR, y = TPR)) + geom_abline(intercept = 0, slope = 1, colour = "gray") + ylab("TPR") + xlab("FPR")
auc
```

### Classify into high/low risk groups

```{r glmnet-cox-3}
metadata <- dataset$testing_metadata
metadata$risk <- as.factor(ifelse(pred > median(pred), "High-risk", "Low-risk"))
km <- survfit(Surv(days_survived, vital_status) ~ risk, data = metadata)
ggsurv(km) + theme_bw() + ggtitle("Kaplan-Meier plot for risk based on combined tsRNA + miRNA signature")
cox <- coxph(Surv(days_survived, vital_status) ~ risk, data = metadata)
summary(cox)
```

### Build new logistic regression model using hazard ratios


```{r glmnet-cox-4}


```

```{r save-session, cache=FALSE}
save.session("survival-features.RSession")
```
