# Process data
_Last modified `r format(Sys.time(), "%I:%M %p on %b %d, %Y")`. This document, R session image, knitr cache, figures, and other associated datasets are located in `cruncher:/inside/grotto/blin/trna-markers/process-reads/`._

```{r setup, echo=FALSE, warning=FALSE, results=FALSE, message=FALSE, errors=FALSE, cache=FALSE}
library(knitr)
opts_chunk$set(cache = TRUE, cache.path = "/inside/grotto/blin/trna-markers/process-reads/cache/process-reads/", eval = TRUE, echo = TRUE, warning = FALSE, results = FALSE, message = FALSE, autodep = TRUE, dev = "png", dpi = 300)
```

```{r libraries, cache=FALSE}
library(stringr)
library(GenomicRanges)
library(GenomicAlignments)
library(BiocParallel)
source('/inside/grotto/blin/programs/convertChrNames.R')
load('/inside/grotto/blin/data/hg19-srnas.RData')
```

This script downloads and remaps reads. 

- v1: Prostate adenocarcinoma (PRAD)
- v2: Lung adenocarcinoma (LUAD) and breast invasive carcinoma (BRCA)
- v3: Hepatocellular carcinoma (LIHC), lower grade glioma (LGG), Lymphoid neoplasm diffuse large b-cell lymphoma (DLBC), colorectal (COAD and READ), pancreatic (PAAD), testicular germ cell (TGCT)
  - acute myeloid leukemia (AML and LAML), glioblastoma (GBM), osteosarcoma (OS) do not have enough samples for miRNA-Seq data.
- v4: Rest of PanCan 12 sample types


## Parse file and clinical metadata

```{r parse-metadata-function}
# metadata_file: sample metadata downloaded from CGHub in tab delimited form
# cancer_code: short abbreviation in lowercase describing the cancer type, e.g. "prad" or "luad"
parseMetadata <- function(metadata_file, cancer_code) {
  metadata <- read.table(metadata_file, header = TRUE, sep = '\t', as.is = TRUE)
  metadata <- metadata[, c('barcode', 'sample_type', 'filename', 'analysis_id', 'analyte_type', 'uploaded', 'participant_id')]
  # convert character "uploaded" column to date object for easier comparison
  metadata$uploaded <- strptime(gsub("/", "-", metadata$uploaded), "%Y-%m-%d")
  # start by getting matched pairs
  # first, check if there actually are normal tissue samples
  parsed_metadata <- data.frame()
  if (any(metadata$sample_type == "NT")) {
  # all NT datasets have at least one TP dataset from the same patient (participant). We can filter out datasets by existence of the NT file, then grab the most recent TP and NT datasets for each patient.
    paired <- subset(metadata, participant_id %in% subset(metadata, sample_type == "NT")$participant_id & participant_id %in% subset(metadata, sample_type == "TP")$participant_id)
    paired$paired <- TRUE # remember which samples are matched tumor/normal
    for (current_id in unique(paired$participant_id)) {
      nt <- subset(paired, participant_id == current_id & sample_type == "NT")
      tp <- subset(paired, participant_id == current_id & sample_type == "TP")
      parsed_metadata <- rbind(parsed_metadata, tp[which(tp$uploaded == max(tp$uploaded)), ], nt[which(nt$uploaded == max(nt$uploaded)), ])
    }
    singles <- metadata[-which(metadata$participant_id %in% parsed_metadata$participant_id), ]
  } else {
    singles <- metadata
  }
  # now for the unpaired samples
  singles$paired <- FALSE
  for (current_id in unique(singles$participant_id)) {
    ps <- subset(singles, participant_id == current_id) # get all samples for this participant
    parsed_metadata <- rbind(parsed_metadata, ps[which(ps$uploaded == max(ps$uploaded)), ])
  }
  # create sample ID numbers and new filenames
  numbering <- 1:nrow(parsed_metadata)
  parsed_metadata$shortids <- numbering
  numbering <- formatC(numbering, flag = "0", digits = 3)
  parsed_metadata$shortnames <- paste0(cancer_code, "-", numbering)
  # reencode upload date as a string - it's encoded as a time object, screwing up data frame operations
  parsed_metadata$uploaded <- as.character(parsed_metadata$uploaded)
  parsed_metadata
}
processClinicalDataFile = function(biotab_file, batch_file) {
  biotab = read.delim(biotab_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  batches = read.table(batch_file, header = TRUE, sep = "\t", as.is = c(FALSE, TRUE))
  biotab$bcr_patient_uuid = tolower(biotab$bcr_patient_uuid) # all data file ids are in lowercase, not sure if matters
  biotab$days_survived = replace(biotab$death_days_to, biotab$death_days_to == "[Not Applicable]", biotab$last_contact_days_to[which(biotab$death_days_to == "[Not Applicable]")]) # either days from initial prognosis to death or initial prognosis to censor time
  biotab$batch = batches[match(biotab$bcr_patient_barcode, substr(batches$barcode, 1, 12)), "batch"]
  biotab
}
```

```{r parse-metadata}
luad_metadata = parseMetadata('/inside/grotto/blin/trna-markers/process-reads/luad-summary.tab', 'luad')
brca_metadata = parseMetadata('/inside/grotto/blin/trna-markers/process-reads/brca-summary.tab', 'brca')
prad_metadata = parseMetadata('/inside/grotto/blin/trna-markers/process-reads/prad-summary.tab', 'prad')
dlbc_metadata = parseMetadata('/inside/grotto/blin/trna-markers/process-reads/dlbc-summary.tab', 'dlbc')
lihc_metadata = parseMetadata('/inside/grotto/blin/trna-markers/process-reads/lihc-summary.tab', 'lihc')
coad_metadata = parseMetadata('/inside/grotto/blin/trna-markers/process-reads/coad-summary.tab', 'coad')
paad_metadata = parseMetadata('/inside/grotto/blin/trna-markers/process-reads/paad-summary.tab', 'paad')
tgct_metadata = parseMetadata('/inside/grotto/blin/trna-markers/process-reads/tgct-summary.tab', 'tgct')
lgg_metadata = parseMetadata('/inside/grotto/blin/trna-markers/process-reads/lgg-summary.tab', 'lgg')
```


## Clinical data

```{r get-clinical-data}
luad_clinical = processClinicalDataFile('luad-clinical-biotab-04032015.tab', 'luad-clinical-batches-04032015.tab')
luad_clinical = with(luad_clinical[match(luad_metadata$participant_id, luad_clinical$bcr_patient_uuid), ], 
                    data.frame(luad_metadata,
                               gender = gender, 
                               race = race,
                               smoker = ifelse(tobacco_smoking_history_indicator == "Lifelong Non-smoker", 0, 1), # NA doesn't return true or false, just NA. this is useful.
                               t_stage = ajcc_tumor_pathologic_pt, 
                               m_stage = ajcc_metastasis_pathologic_pm, 
                               n_stage = ajcc_nodes_pathologic_pn,
                               stage = ajcc_pathologic_tumor_stage,
                               days_survived = as.integer(days_survived), 
                               vital_status = ifelse(vital_status == "Alive", 0, 1),
                               age = age_at_initial_pathologic_diagnosis,
                               batch = batch))
luad_metadata = luad_clinical
brca_clinical = processClinicalDataFile('brca-clinical-biotab-04032015.tab', 'brca-clinical-batches-04032015.tab')
brca_clinical = with(brca_clinical[match(brca_metadata$participant_id, brca_clinical$bcr_patient_uuid), ], 
                     data.frame(brca_metadata,
                                ethnicity = ethnicity,
                                race = race,
                                vital_status = ifelse(vital_status == "Alive", 0, 1),
                                days_survived = as.integer(days_survived),
                                age = age_at_diagnosis,
                                t_stage = ajcc_tumor_pathologic_pt, 
                                m_stage = ajcc_metastasis_pathologic_pm, 
                                n_stage = ajcc_nodes_pathologic_pn,
                                stage = ajcc_pathologic_tumor_stage,
                                er_status = er_status_by_ihc,
                                pr_status = pr_status_by_ihc,
                                her2_status = her2_status_by_ihc,
                                batch = batch))
brca_metadata = brca_clinical
prad_clinical = processClinicalDataFile('prad-clinical-biotab-04012015.tab', 'prad-clinical-batches-04012015.tab')
prad_clinical = with(prad_clinical[match(prad_metadata$participant_id, prad_clinical$bcr_patient_uuid), ],
                     data.frame(prad_metadata,
                                days_survived = as.integer(days_survived),
                                vital_status = ifelse(vital_status == "Alive", 0, 1),
                                t_stage = pathologic_T,
                                m_stage = pathologic_M,
                                n_stage = pathologic_N,
                                gleason_score = as.integer(gleason_score),
                                age = abs(round(as.integer(birth_days_to)/365)),
                                batch = batch)) # maybe change to age_at _initial_pathologic_diagnosis? 
prad_metadata = prad_clinical
lihc_clinical = processClinicalDataFile('lihc-clinical-biotab-06062016.tab', 'lihc-clinical-batches-06062016.tab') 
lihc_clinical = with(lihc_clinical[match(lihc_metadata$participant_id, lihc_clinical$bcr_patient_uuid), ],
                     data.frame(lihc_metadata,
                                days_survived = as.integer(days_survived),
                                vital_status = ifelse(vital_status == "Alive", 0, 1),
                                t_stage = pathologic_T,
                                m_stage = pathologic_M,
                                n_stage = pathologic_N,
                                hepatitis_b = str_detect(history_hepato_carcinoma_risk_factors, "Hepatitis B"),
                                hepatitis_c = str_detect(history_hepato_carcinoma_risk_factors, "Hepatitis C"),
                                alcoholic = str_detect(history_hepato_carcinoma_risk_factors, "Alcohol consumption"),
                                nonalcoholic_predisposition = str_detect(history_hepato_carcinoma_risk_factors, "Non-Alcoholic"),
                                history_hepato_carcinoma_risk_factors = history_hepato_carcinoma_risk_factors,
                                batch = batch))
lihc_metadata = lihc_clinical
lgg_clinical = processClinicalDataFile('lgg-clinical-biotab-06062016.tab', 'lgg-clinical-batches-06062016.tab')
lgg_clinical = with(lgg_clinical[match(lgg_metadata$participant_id, lgg_clinical$bcr_patient_uuid), ],
                    data.frame(lgg_metadata,
                               days_survived = as.integer(days_survived),
                               vital_status = ifelse(vital_status == "Alive", 0, 1),
                               tumor_grade = tumor_grade,
                               history_headaches = history_headaches,
                               history_seizures = history_seizures,
                               idh1 = replace(idh1_mutation_found, idh1_mutation_found == "[Not Applicable]", idh1_mutation_test_indicator[which(idh1_mutation_found == "[Not Applicable]")]),
                               batch = batch))
lgg_metadata = lgg_clinical
```



```{r download-files}
metadatas <- list(luad = luad_metadata, brca = brca_metadata, prad = prad_metadata, dlbc = dlbc_metadata, lihc = lihc_metadata, coad = coad_metadata, paad = paad_metadata, tgct = tgct_metadata, lgg = lgg_metadata)
currentwd <- getwd()
setwd('/inside/grotto/blin/trna-markers/datasets/raw')
for (cancer_code in names(metadatas)) { 
  metadata <- metadatas[[cancer_code]]
  for (analysis_id in metadata$analysis_id) {
  	if (file.exists(analysis_id)) next # if exists, assume file was downloaded correctly and is not corrupted
    system(paste0('echo gtdownload -c $CGKEY -d ', analysis_id, cancer_code))
    system(paste0('gtdownload -c $CGKEY -d ', analysis_id))
  }
  file.symlink(paste0(getwd(), '/', metadata$analysis_id, '/', metadata$barcode, '_mirna.bam'), paste0('/inside/grotto/blin/trna-markers/datasets/', cancer_code, '/mirna/', metadata$shortnames, '-raw.bam'))
}
setwd(currentwd)
```

We are missing `01810b1a-84e4-43d5-8a1e-42b132a1126f`, which is `brca-0939-TP.bam`. It's been removed for now.

## Remap reads

To count both tRFs processed from mature and immature tRNAs, we need to map to both hg19 and mature tRNAs.

### hg19

Andrew has a few settings that I've incorporated too, with his reasoning:
  - --very-sensitive: necessary due to mismatch caused by modified base misreads 
  - --np 5 --ignore-quals: quality scores are irrelevant, and N should be scored as only slightly better than a mismatch. ignoring quals forces the mismatches to have a single score, which is necessary for precise N weighting.

```{r remap, engine='bash', results='hide'}
remap() {
  currentwd=`pwd`
  cd $1
  for prefix in `ls *-raw.bam | cut -f 1,2 -d -`; do
    if [ ! -e $prefix.bam ]
    then
      if [ ! -e $prefix.fastq ]
      then
        bam2fastq --force -o $prefix-raw# `readlink $prefix-raw.bam`
        mv ${prefix}-raw_M $prefix.fastq
        rm ${prefix}-raw_1 ${prefix}-raw_2
      fi
      bowtie2 -x ~/grotto/data/hg19-tRNAgenome -k 100 --very-sensitive --ignore-quals --np 5 -p 32 -U ${prefix}.fastq -S $prefix-temp-1.sam
      samtools view -S -F 4 -h -u $prefix-temp-1.sam | samtools view -F 512 -h  - > $prefix-temp-2.sam # filter out untemp and low quality reads
      bowtie2-best-mapped.py $prefix-temp-2.sam | samtools view -S -u - | samtools sort - $prefix # auto appends .bam
    fi
    rm -f $prefix.fastq
    rm -f $prefix-temp-1.sam
    rm -f $prefix-temp-2.sam
  done
  cd $currentwd
}
remap /inside/grotto/blin/trna-markers/datasets/luad/mirna
remap /inside/grotto/blin/trna-markers/datasets/brca/mirna
remap /inside/grotto/blin/trna-markers/datasets/prad/mirna
#remap /inside/grotto/blin/trna-markers/datasets/dlbc/mirna
remap /inside/grotto/blin/trna-markers/datasets/lihc/mirna
#remap /inside/grotto/blin/trna-markers/datasets/coad/mirna
#remap /inside/grotto/blin/trna-markers/datasets/paad/mirna
#remap /inside/grotto/blin/trna-markers/datasets/tgct/mirna
remap /inside/grotto/blin/trna-markers/datasets/lgg/mirna
```

## Count reads

There's a few types of sRNAs. We can get most of them from the hg19 mappings, but for tRF-1, tRF-3, and mature tRNA halves, we need the mappings to mature tRNAs.

So here's each sRNA, by source and destination.

- tRF-1: hg19/Trailers from Andrew's tRNA half db
- immature tRNA 5' half: hg19/Trailers from Andrew's tRNA half db
- mature tRNA 3' half/tRF3: tRNAs/split from anticodon
- mature tRNA 5' half/tRF5: tRNAs/split from anticodon
- miRNAs: hg19/miRbase
- snoRNAs: hg19/DASHR

```{r count-features-function}
countFeaturesFromBam = function(metadata, features, dir = ".") {
  metadata = metadata[which(file.exists(paste0(dir, "/", metadata$shortnames, ".bam"))), ]
  trf1s_plus = features[features$class == "tRF-1" & strand(features) == "+"]
  trf1s_minus = features[features$class == "tRF-1" & strand(features) == "-"]
  trf3s = features[features$class == "tRF-3"]
  trf5s = features[features$class == "tRF-5"]
  features = features[!(features$class %in% c("tRF-1", "tRF-3", "tRF-5"))]
  # use a for loop because r will run out of memory
  counts = mclapply(1:nrow(metadata), function(i) {
    row = metadata[i, ]
    system(paste("echo Counting sRNAs from", row$shortnames))
    bamfile = paste0(dir, "/", row$shortnames, ".bam")
    sample = readGAlignments(bamfile)
    sample = convertChrNames(sample, "Ensembl")
    sample_counts = rbind(data.frame(counts = countOverlaps(features, sample, minoverlap = 10)),
                          data.frame(counts = countOverlaps(trf1s_plus, sample, minoverlap = 10, type = 'start')),
                          data.frame(counts = countOverlaps(trf1s_minus, sample, minoverlap = 10, type = 'end')),
                          data.frame(counts = countOverlaps(trf3s, sample, minoverlap = 10, type = 'end', maxgap = 3)), # in case of a CCA tail. We don't have to worry about aligning to 3 past the gap since these are the mature tRNAs.
                          data.frame(counts = countOverlaps(trf5s, sample, minoverlap = 10, type = 'start')))
    colnames(sample_counts) = row$barcode
    sample_counts
  }, mc.cores = 20)
  names = unlist(lapply(counts, colnames))
  counts = as.data.frame(matrix(unlist(counts), ncol = nrow(metadata)))
  rownames(counts) = c(features$tx_name, trf1s_plus$tx_name, trf1s_minus$tx_name, trf3s$tx_name, trf5s$tx_name)
  colnames(counts) = names
  counts
}
```

```{r count-features-luad}
luad_raw_counts <- countFeaturesFromBam(luad_metadata, srnas, '/inside/grotto/blin/trna-markers/datasets/luad/mirna/')
save(file = 'luad-raw-counts.RData', luad_raw_counts)
save(file = 'luad-metadata.RData', luad_metadata) 
```

```{r count-features-brca}
brca_raw_counts <- countFeaturesFromBam(brca_metadata, srnas, '/inside/grotto/blin/trna-markers/datasets/brca/mirna/')
save(file = 'brca-raw-counts.RData', brca_raw_counts)
save(file = 'brca-metadata.RData', brca_metadata)
```

```{r count-features-prad}
prad_raw_counts <- countFeaturesFromBam(prad_metadata, srnas, '/inside/grotto/blin/trna-markers/datasets/prad/mirna/')
save(file = 'prad-raw-counts.RData', prad_raw_counts)
save(file = 'prad-metadata.RData', prad_metadata)
```

```{r count-features-lihc}
lihc_raw_counts <- countFeaturesFromBam(lihc_metadata, srnas, '/inside/grotto/blin/trna-markers/datasets/lihc/mirna/')
save(file = 'lihc-raw-counts.RData', lihc_raw_counts)
save(file = 'lihc-metadata.RData', lihc_metadata)
```

```{r count-features-lgg}
lgg_raw_counts <- countFeaturesFromBam(lgg_metadata, srnas, '/inside/grotto/blin/trna-markers/datasets/lgg/mirna/')
save(file = 'lgg-raw-counts.RData', lgg_raw_counts)
save(file = 'lgg-metadata.RData', lgg_metadata)
```

```{r save-session, cache=FALSE}
save.session("process-reads.RSession")
```
