# trials
_Last modified `r format(Sys.time(), "%I:%M %p on %b %d, %Y")`. This document, R session image, version history, knitr cache, figures, and other associated datasets are located in `/inside/grotto/blin/trna-markers/luad/elastic-nets/`._

```{r setup, echo=FALSE, warning=FALSE, results=FALSE, message=FALSE, errors=FALSE, cache=FALSE}
library(knitr)
opts_chunk$set(cache = TRUE, cache.path = "/inside/grotto/blin/trna-markers/luad/elastic-nets/trials-cache/", eval = TRUE, echo = TRUE, warning = FALSE, results = FALSE, message = FALSE, autodep = TRUE, dev = "png", dpi = 300)
```

```{r libraries, cache=FALSE}
library(gelnet)
library(glmnet)
library(ROCR)
library(GenomicRanges)
library(plyr)
library(stringr)
set.seed(12)
```

```{r load-data}
load('/inside/home/blin/grotto/data/hg19-srnas.RData')
features <- unique(srnas[srnas$class %in% c("trf-1", "trf-3", "trf-5", "actRF-3", "actRF-5", "trailer")]$tx_name)
load('/inside/home/blin/grotto/trna-markers/process-reads/luad-counts.RData')
luad_clinical <- luad_clinical[match(colnames(luad_adjusted_counts), luad_clinical$barcode), ]
```

## Sample data

We need to create 100 random datasets with training and test data. Each dataset will contain an equal amount of paired data.


```{r sampling-function}
randomSamplePatients <- function(metadata) {
	# sample pairs of samples first
	paired_metadata <- metadata[metadata$paired, ]
	paired_samples <- sample(as.factor(paired_metadata$participant_id), nrow(paired_metadata)/2)
	# then sample single sample patients
	unpaired_metadata <- metadata[!metadata$paired, ]
	unpaired_samples <- sample(as.factor(unpaired_metadata$participant_id), nrow(unpaired_metadata)/2)
	# combine patient ids
	c(as.character(paired_samples), as.character(unpaired_samples))
}
setupTrainingTestingSets <- function(metadata, counts) {
  random_sample <- randomSamplePatients(metadata)
  training_counts <-  counts[, metadata$participant_id %in% random_sample]
  training_metadata <- metadata[metadata$participant_id %in% random_sample, ]
  testing_counts <- counts[, !(metadata$participant_id %in% random_sample)]
  testing_metadata <- metadata[!(metadata$participant_id %in% random_sample), ]
  list(training_counts = training_counts, training_metadata = training_metadata, testing_counts = testing_counts, testing_metadata = testing_metadata)
}
```

```{r sampling, eval=FALSE}
ntrials <- 100
luad_datasets <- lapply(1:ntrials, function(i) {
  if (i %% 10 == 0) system(paste("echo", i, "/", ntrials, "trials"))
  setupTrainingTestingSets(luad_clinical, luad_adjusted_counts)
})
save(file = "luad-datasets.RData", luad_datasets)
```

## Build elastic nets

```{r elastic-nets-function}
buildTestElasticNet <- function(dataset, features, randomize = FALSE) {
  training_counts <- t(dataset$training_counts[rownames(dataset$training_counts) %in% features, ])
  training_response <- dataset$training_metadata[match(colnames(dataset$training_counts), dataset$training_metadata$barcode), ]$sample_type
  testing_counts <- t(dataset$testing_counts[rownames(dataset$testing_counts) %in% features, ])
  testing_response <- dataset$testing_metadata[match(colnames(dataset$testing_counts), dataset$testing_metadata$barcode), ]$sample_type
  model <- glmnet(training_counts, training_response, family = "binomial", alpha = 0.5)
  chosen_model <- which.min(abs(model$df - 20)) # select for ~ 20 features)
  pred <- predict(model, newx = testing_counts, s = model$lambda[chosen_model])
  prob <- prediction(pred, testing_response)
  if (randomize) prob <- prediction(pred, sample(testing_response))
  perf <- performance(prob, "tpr", "fpr")
  roc <- data.frame(TPR = unlist(perf@y.values), FPR = unlist(perf@x.values))
  coef <- model$beta[, chosen_model]
  coef <- coef[coef != 0]
  auc <- unlist(performance(prob, "auc")@y.values)
  auc <- round(auc, 3)
  list(roc = roc, coef = coef, auc = auc)
}
```

```{r randomize-performance, fig.width=10, fig.height=4}
luad_tsrna_elnets <- lapply(luad_datasets, buildTestElasticNet, features, randomize = TRUE)
luad_mirna_elnets <- lapply(luad_datasets, buildTestElasticNet, unique(srnas[str_detect(srnas$class, "mi")]$tx_name), randomize = TRUE)
df <- ldply(1:length(luad_tsrna_elnets), function(i) data.frame(luad_tsrna_elnets[[i]]$roc, Trial = as.factor(paste0(i, " (AUC = ", luad_tsrna_elnets[[i]]$auc, ")")), Class = "tsRNA"))
df <- rbind(df, ldply(1:length(luad_mirna_elnets), function(i) data.frame(luad_mirna_elnets[[i]]$roc, Trial = as.factor(paste0(i, " (AUC = ", luad_mirna_elnets[[i]]$auc, ")")), Class = "miRNA")))
ggplot(df) + geom_line(aes(x = FPR, y = TPR, color = Trial), alpha = 0.3) + geom_abline(intercept = 0, slope = 1, colour = "gray") + ylab("TPR") + xlab("FPR") + facet_wrap(~ Class) + guides(color = FALSE)
```


```{r elnet-performance}
luad_tsrna_elnets <- lapply(luad_datasets, buildTestElasticNet, features)
luad_mirna_elnets <- lapply(luad_datasets, buildTestElasticNet, unique(srnas[str_detect(srnas$class, "mi")]$tx_name))
luad_snorna_elnets <- lapply(luad_datasets, buildTestElasticNet, unique(srnas[srnas$class == "snoRNA"]$tx_name))
luad_pirna_elnets <- lapply(luad_datasets, buildTestElasticNet, unique(srnas[srnas$class == "piRNA"]$tx_name))
luad_half_elnets <- lapply(luad_datasets, buildTestElasticNet, unique(srnas[srnas$class %in% c("fivehalf", "threehalf")]$tx_name))
luad_trna_elnets <- lapply(luad_datasets, buildTestElasticNet, unique(srnas[srnas$class == "tRNA"]$tx_name))
```

```{r plot-elnet-performance, fig.width=15, fig.height=8}
df <- ldply(1:length(luad_tsrna_elnets), function(i) data.frame(luad_tsrna_elnets[[i]]$roc, Trial = as.factor(paste0(i, " (AUC = ", luad_tsrna_elnets[[i]]$auc, ")")), Class = "tsRNA"))
df <- rbind(df, ldply(1:length(luad_mirna_elnets), function(i) data.frame(luad_mirna_elnets[[i]]$roc, Trial = as.factor(paste0(i, " (AUC = ", luad_mirna_elnets[[i]]$auc, ")")), Class = "miRNA")))
df <- rbind(df, ldply(1:length(luad_snorna_elnets), function(i) data.frame(luad_snorna_elnets[[i]]$roc, Trial = as.factor(paste0(i, " (AUC = ", luad_snorna_elnets[[i]]$auc, ")")), Class = "snoRNA")))
df <- rbind(df, ldply(1:length(luad_pirna_elnets), function(i) data.frame(luad_pirna_elnets[[i]]$roc, Trial = as.factor(paste0(i, " (AUC = ", luad_pirna_elnets[[i]]$auc, ")")), Class = "piRNA")))
df <- rbind(df, ldply(1:length(luad_half_elnets), function(i) data.frame(luad_half_elnets[[i]]$roc, Trial = as.factor(paste0(i, " (AUC = ", luad_half_elnets[[i]]$auc, ")")), Class = "tRNA half")))
df <- rbind(df, ldply(1:length(luad_trna_elnets), function(i) data.frame(luad_trna_elnets[[i]]$roc, Trial = as.factor(paste0(i, " (AUC = ", luad_trna_elnets[[i]]$auc, ")")), Class = "tRNA")))
ggplot(df) + geom_line(aes(x = FPR, y = TPR, color = Trial), alpha = 0.3) + geom_abline(intercept = 0, slope = 1, colour = "gray") + ylab("TPR") + xlab("FPR") + facet_wrap(~ Class, ncol = 3) + guides(color = FALSE)
```

```{r}
mapply(function(elnets, name) paste(mean(unlist(lapply(1:length(elnets), function(i) elnets[[i]]$auc))), name), elnets = list(luad_tsrna_elnets, luad_mirna_elnets, luad_snorna_elnets, luad_pirna_elnets, luad_half_elnets, luad_trna_elnets), name = c("tsrna", "mirna", "snorna", "pirna", "half", "trna"))
```


```{r save-session, eval=FALSE}
save.session("trials.RSession")
```
