# luad-survival
_Last modified `r format(Sys.time(), "%I:%M %p on %b %d, %Y")`. This document, R session image, knitr cache, figures, and other associated datasets are located in `cruncher:/inside/grotto/blin/trna-markers/luad/luad-survival/`._

```{r setup, echo=FALSE, warning=FALSE, results=FALSE, message=FALSE, errors=FALSE, cache=FALSE}
library(knitr)
opts_chunk$set(cache = TRUE, cache.path = "/inside/grotto/blin/trna-markers/luad/luad-survival/cache/luad-survival/", eval = TRUE, echo = TRUE, warning = FALSE, results = FALSE, message = FALSE, autodep = TRUE, dev = "png", dpi = 300)
```

```{r libraries, cache=FALSE}
library(survival)
library(glmnet)
library(GenomicRanges)
library("RColorBrewer")
library(pheatmap)
source('/inside/home/blin/lib/R/ggsurv.R')
source('/inside/grotto/blin/trna-markers/luad/predict/predict.R')
load('/inside/home/blin/grotto/data/hg19-srnas.RData')
load('/inside/home/blin/grotto/trna-markers/process-reads/luad-counts.RData')
load('/inside/home/blin/grotto/trna-markers/process-reads/luad-metadata.RData')
luad_metadata = luad_metadata[match(colnames(luad_adjusted_counts), luad_metadata$barcode), ]
```

## Univariate survival analysis for clinical covariates

First, we'll independently test each covariate and feature to see what is implicated in survival.

### Subtype

```{r univariate-subtype}
km_metadata = na.omit(luad_metadata[luad_metadata$sample_type != "NT" & !is.na(luad_metadata$days_survived), ]) # get rid of NAs, remove normal samples (e.g. duplicates), and put the time period as 1 month - 8 years
load('/inside/grotto/blin/trna-markers/luad/cluster-subtypes/clusters.RData')
clusters = clusters[clusters$known, ]
metadata = km_metadata[substr(km_metadata$barcode, 1, 12) %in% clusters$barcode, ]
metadata$subtype = as.factor(clusters$subtype[match(substr(metadata$barcode, 1, 12), clusters$barcode)])
coxph(Surv(days_survived, vital_status) ~ subtype, data = metadata)
```

### Tumor stage

```{r univariate-stage}
metadata = km_metadata[km_metadata$stage != "[Not Available]", ]
levels(metadata$stage) = list(I = c("Stage I", "Stage IA", "Stage IB"), II = c("Stage II", "Stage IIA", "Stage IIB"), III = c("Stage III", "Stage IIIA", "Stage IIIB"), IV = "Stage IV")
km = survfit(Surv(days_survived, vital_status) ~ stage, data = metadata)
ggsurv(km) + theme_bw() + ggtitle("Kaplan-Meier plot grouped by stage")
coxph(Surv(days_survived, vital_status) ~ stage, data = metadata)
```

### Lymph node involvement

```{r univariate-lymph}
metadata = km_metadata[km_metadata$n_stage %in% c("N0", "N1", "N2", "N3"), ]
metadata$n_stage = droplevels(metadata$n_stage)
km = survfit(Surv(days_survived, vital_status) ~ n_stage, data = metadata)
ggsurv(km) + theme_bw() + ggtitle("Kaplan-Meier plot grouped by lymph node spread")
coxph(Surv(days_survived, vital_status) ~ n_stage, data = metadata)
```

### Metastasis

```{r univariate-metastasis}
metadata = km_metadata[km_metadata$m_stage %in% c("M0", "M1", "M1a", "M1b"), ]
levels(metadata$m_stage) = list(M0 = "M0", M1 = c("M1", "M1a", "M1b"))
km = survfit(Surv(days_survived, vital_status) ~ m_stage, data = metadata)
ggsurv(km) + theme_bw() + ggtitle("Kaplan-Meier plot grouped by metastasis")
summary(coxph(Surv(days_survived, vital_status) ~ m_stage, data = metadata))
```

### Smoker vs non-smoker

```{r univariate-smoking}
coxph(Surv(days_survived, vital_status) ~ smoker, data = km_metadata)
```

### Gender

```{r univariate-gender}
coxph(Surv(days_survived, vital_status) ~ gender, data = km_metadata)
```

### Race
```{r univariate-race}
metadata = km_metadata
levels(metadata$race) = list(White = "WHITE", Black = "BLACK OR AFRICAN AMERICAN", Asian = "ASIAN", Unknown = c("[Not Available", "[Not Evaluated]", "[Unknown]"))
km = survfit(Surv(days_survived, vital_status) ~ race, data = metadata)
coxph(Surv(days_survived, vital_status) ~ race, data = metadata)
```

### Age

```{r univariate-age}
metadata = km_metadata
metadata$age = ifelse(as.integer(as.character(metadata$age)) <= 65, "<= 65", "> 65")
km = survfit(Surv(days_survived, vital_status) ~ age, data = metadata)
coxph(Surv(days_survived, vital_status) ~ age, data = metadata)
```

Out of all of these, only lymph node involvement and cancer stage are significant predictors of survival.

## Univariate survival analyses for features

```{r univariate-features}
metadata = luad_metadata[luad_metadata$sample_type == "TP" & !is.na(luad_metadata$days_survived) & luad_metadata$days_survived > 0, ]
counts = luad_adjusted_counts[rownames(luad_adjusted_counts) %in% srnas[srnas$class %in% c("tRF-1", "tRF-3", "tRF-5", "miRNA")]$tx_name, colnames(luad_adjusted_counts) %in% metadata$barcode]
metadata$dummy = as.factor(1)
dataset = setupTrainingTestingSets(metadata, "dummy", counts)
features = rownames(counts)
hazard_ratios = data.frame(t(sapply(features, function(feature) {
  training_counts = t(dataset$training_counts)[, feature]
  training_metadata = cbind(dataset$training_metadata, feature = training_counts)
  cox = summary(coxph(Surv(days_survived, vital_status) ~ feature, data = training_metadata))
  hazard = cox$conf.int[1]
  conf_int = cox$conf.int[3:4]
  pval = coef(cox)[5]
  c(hazard, conf_int, pval)
  })))
colnames(hazard_ratios) = c('hazard', 'lower', 'upper', 'pvalue')
```

We have `r sum(hazard_ratios$pvalue <= 0.05)` features that are significant (p < 0.05), with `r sum(hazard_ratios$pvalue[str_detect(rownames(hazard_ratios), "hsa-mir|hsa-let")] <= 0.05)` miRNAs and `r sum(hazard_ratios$pvalue[str_detect(rownames(hazard_ratios), "tRF|trailer")] <= 0.05)` tRNA fragments. 


## Identify signatures


```{r univariate-features-2}
mirna_signature = hazard_ratios[str_detect(rownames(hazard_ratios), "hsa-mir|hsa-let"), ]
mirna_signature = mirna_signature[order(mirna_signature$pvalue), ][1:20, ]
kable(mirna_signature)
tsrna_signature = hazard_ratios[str_detect(rownames(hazard_ratios), "tRF|trailer"), ]
tsrna_signature = tsrna_signature[order(tsrna_signature$pvalue), ][1:20, ]
kable(tsrna_signature)
srna_signature = hazard_ratios[order(hazard_ratios$pvalue), ][1:20, ]
kable(srna_signature)
```

## Train signatures

```{r train-signatures-function}
trainTestSignatures = function(dataset, features) {
  # build model
  training_counts = t(dataset$training_counts[rownames(dataset$training_counts) %in% features, ])
  training_response = Surv(dataset$training_metadata$days_survived, dataset$training_metadata$vital_status)
  model = cv.glmnet(training_counts, training_response, alpha = 1, nlambda = 250, family = "cox")
  plot(model) # plot lambdas
  plot(model$glmnet.fit, label = TRUE) # plot coefficient shrinkage

  # test model - predict 3 year survival rate on testing data
  testing_metadata = dataset$testing_metadata
  testing_counts = t(dataset$testing_counts[rownames(dataset$training_counts) %in% features, colnames(dataset$testing_counts) %in% testing_metadata$barcode])
  testing_response = as.factor(ifelse(testing_metadata$days_survived >= 1095, ">3", "<3")) # test 3 year survival rate
  pred = predict(model, newx = testing_counts, s = model$lambda.min)
  prob = prediction(-pred, testing_response) # flip signs because we trained with survival data, and a positive coefficient gives a higher risk of death
  perf = performance(prob, "tpr", "fpr")
  roc = data.frame(TPR = unlist(perf@y.values), FPR = unlist(perf@x.values))
  auc = unlist(performance(prob, "auc")@y.values)
  auc = round(auc, 3)

  # classify into high/low risk sets
  testing_metadata$risk = as.factor(ifelse(pred > median(pred), "High-risk", "Low-risk"))
  km = survfit(Surv(days_survived, vital_status) ~ risk, data = testing_metadata)
  km = ggsurv(km) + theme_bw() + ggtitle("Kaplan-Meier plot for risk based on signature")
  cox = coxph(Surv(days_survived, vital_status) ~ risk, data = testing_metadata)

  list(model = model, coef = coef(model, s = model$lambda.min), score = pred[, ], roc = roc, auc = auc, cox = summary(cox), km = km)
}

```

To get a sense of the data, let's look at all of the clinical covariates:

```{r dataset-covariates}
c(table(dataset$training_metadata$stage), table(dataset$testing_metadata$stage))
c(table(dataset$training_metadata$m_stage), table(dataset$testing_metadata$m_stage))
c(table(dataset$training_metadata$n_stage), table(dataset$testing_metadata$n_stage))
c(table(ifelse(dataset$training_metadata$age > 65, ">65", "<=65")), table(ifelse(dataset$testing_metadata$age > 65, ">65", "<=65")))
c(table(dataset$training_metadata$gender), table(dataset$testing_metadata$gender))
c(table(dataset$training_metadata$smoker), table(dataset$testing_metadata$smoker))
c(table(dataset$training_metadata$race), table(dataset$testing_metadata$race))
```

### tsRNA signature performance

```{r train-signatures-tsrna}
tsrna = trainTestSignatures(dataset, rownames(tsrna_signature))
tsrna$coef
coef(tsrna$cox)
tsrna$km + ggtitle("")
```

### microRNA signature performance

```{r train-signatures-mirna}
mirna = trainTestSignatures(dataset, rownames(mirna_signature))
mirna$coef
coef(mirna$cox)
mirna$km + ggtitle("")
```

### Combined signature performance

```{r train-signatures-srna}
srna = trainTestSignatures(dataset, rownames(srna_signature))
srna$coef
coef(srna$cox)
srna$km + ggtitle("")
```

### Predicting 3-year survival

```{r predict-survival}
roc = rbind(cbind(tsrna$roc, Class = "tsRNA"), cbind(mirna$roc, Class = "miRNA"), cbind(srna$roc, Class = "Combined"))
ggplot(roc) + geom_line(aes(x = FPR, y = TPR, color = Class)) + geom_abline(intercept = 0, slope = 1, colour = "gray") + ylab("TPR") + xlab("FPR") + theme_bw()
c(tsRNA = tsrna$auc, miRNA = mirna$auc, sRNA = srna$auc)
```

## Multivariate survival analyses

Let's test whether this sRNA signature has prognostic value independent of stage and lymph node involvement. First, the univariate estimate:

```{r multivariate-cox-1}
metadata = dataset$testing_metadata
metadata$score = srna$score
cox = coxph(Surv(days_survived, vital_status) ~ score, data = metadata)
kable(data.frame(Coef = cox$coef, Hazard = summary(cox)$conf.int[1], Lower = summary(cox)$conf.int[3], Upper = summary(cox)$conf.int[4], Pval = summary(cox)$coef[5]))
```

Then, let's see if stage is a confounding factor. If it is, then the confidence intervals should shrink (added precision), and the hazard ratio should appreciably change.

```{r multivariate-cox-2}
metadata = metadata[metadata$stage != "[Not Available]", ]
levels(metadata$stage) = list(I = c("Stage I", "Stage IA", "Stage IB"), II = c("Stage II", "Stage IIA", "Stage IIB"), III = c("Stage III", "Stage IIIA", "Stage IIIB"), IV = "Stage IV")
cox = coxph(Surv(days_survived, vital_status) ~ score + stage, data = metadata)
kable(data.frame(Coef = cox$coef, Hazard = summary(cox)$conf.int[, 1], Lower = summary(cox)$conf.int[, 3], Upper = summary(cox)$conf.int[, 4], Pval = summary(cox)$coef[, 5]))
```

The confidence interval actually increases when we add cancer stage as a covariate, so it seems independent. But there's some conflicting evidence. The coefficient shrank by a factor of 7%. 

By the way, since stage is a categorical variable, ANOVA can see whether stage is a significant covariate _after_ adjustment for the effect of risk score. It's significant.

```{r multivariate-cox-3}
anova(cox)
```

Now, we test if there's an interaction between score and stage. 

```{r multivariate-cox-4}
cox = coxph(Surv(days_survived, vital_status) ~ score + stage + score:stage, data = metadata)
kable(data.frame(Coef = cox$coef, Hazard = summary(cox)$conf.int[, 1], Lower = summary(cox)$conf.int[, 3], Upper = summary(cox)$conf.int[, 4], Pval = summary(cox)$coef[, 5]))
```

The interaction term is not significant, so we can safely say there's no interaction.

Same analysis for lymph node spread.

```{r multivariate-cox-5}
metadata = dataset$testing_metadata
metadata$score = tsrna$score
metadata = metadata[metadata$n_stage %in% c("N0", "N1", "N2", "N3"), ]
metadata$n_stage = droplevels(metadata$n_stage)
cox = coxph(Surv(days_survived, vital_status) ~ score + n_stage, data = metadata)
kable(data.frame(Coef = cox$coef, Hazard = summary(cox)$conf.int[, 1], Lower = summary(cox)$conf.int[, 3], Upper = summary(cox)$conf.int[, 4], Pval = summary(cox)$coef[, 5]))
anova(cox)
cox = coxph(Surv(days_survived, vital_status) ~ score + n_stage + score:n_stage, data = metadata)
kable(data.frame(Coef = cox$coef, Hazard = summary(cox)$conf.int[, 1], Lower = summary(cox)$conf.int[, 3], Upper = summary(cox)$conf.int[, 4], Pval = summary(cox)$coef[, 5]))
```

Similar considerations as above. Hazard ratio doesn't change much, and confidence intervals widen. As a whole, lymph node spread is significant after adjusting for risk score. Interaction terms fail to reach significance.

## Visualizations

### Risk scores for sRNA signature

```{r risk-score-plot, fig.width=12, fig.height=3}
df = data.frame(Sample = as.character(dataset$testing_metadata$barcode), Score = srna$score, stringsAsFactors = FALSE)
df = df[order(df$Score), ]
df$Sample = factor(df$Sample, levels = df$Sample)
ggplot(df) + geom_point(aes(x = Sample, y = Score)) + theme(axis.text.x = element_blank())
```

### Expression heatmaps

```{r srna-heatmap-1, fig.width=8, fig.height=5}
df = log1p(as.matrix(luad_adjusted_counts[rownames(luad_adjusted_counts) %in% rownames(hazard_ratios[rownames(hazard_ratios) %in% rownames(srna_signature), ]), ]))
heatmap_metadata = luad_metadata[luad_metadata$barcode %in% colnames(df), ]
# subtype
load('/inside/grotto/blin/trna-markers/luad/cluster-subtypes/clusters.RData')
clusters$subtype[!clusters$known] = "Unknown"
heatmap_metadata$subtype = as.factor(clusters$subtype[match(substr(heatmap_metadata$barcode, 1, 12), clusters$barcode)])
# stage
levels(heatmap_metadata$stage) = list(I = c("Stage I", "Stage IA", "Stage IB"), II = c("Stage II", "Stage IIA", "Stage IIB"), III = c("Stage III", "Stage IIIA", "Stage IIIB"), IV = "Stage IV", Unknown = '[Not Available]')
# n stage
levels(heatmap_metadata$n_stage) = list(N0 = "N0", N1 = "N1", N2 = "N2", N3 = "N3", Unknown = c("NX", "[Not Available]"))
# smoker
heatmap_metadata$smoker = as.factor(heatmap_metadata$smoker)
levels(heatmap_metadata$smoker) = list(Smoker = 1, Nonsmoker = 0)
# heatmap annotation
annot_col = data.frame(stage = heatmap_metadata$stage, subtype = heatmap_metadata$subtype, n_stage = heatmap_metadata$n_stage, smoker = heatmap_metadata$smoker)
rownames(annot_col) = colnames(df)
annot_row = data.frame(sRNA = as.factor(srnas[match(rownames(df), srnas$tx_name), ]$class))
rownames(annot_row) = rownames(df)
annot_colors = list(stage = setNames(c(brewer.pal(4, "Set1"), "grey"), levels(heatmap_metadata$stage)),
                    subtype = setNames(c(brewer.pal(3, "Set2"), "grey"), levels(heatmap_metadata$subtype)),
                    n_stage = setNames(c(brewer.pal(4, "Set3"), "grey"), levels(heatmap_metadata$n_stage)),
                    smoker = setNames(c("#8dd3c7", "#ffffb3", "grey"), levels(heatmap_metadata$smoker)),
                    sRNA = setNames(c("#fc9272", "#de2d26", "#08519c"), levels(annot_row$sRNA)))
pheatmap(df, annotation_col = annot_col, annotation_row = annot_row, annotation_colors = annot_colors, fontsize = 8,  show_colnames = FALSE, show_rownames = FALSE, trace = "none", color = brewer.pal(10, "Reds")) 
```


```{r save-session, echo=FALSE, results='hide', cache=FALSE}
save.session("luad-survival.RSession")
```
